---
title: "Predicting the BTC halving"
author: "Charliemarketplace"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "styles.css"
    includes:
      in_header: header.html
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---


```{r, message=FALSE, warning=FALSE}
library(reactable)
library(dplyr)
library(plotly)
```

# Concept 

The Bitcoin halving occurs every 210,000 blocks. A Bitcoin block occurs *roughly* 
every 10 minutes, but can vary based on hashrate and difficulty adjustments. 

Using just on chain data, can we forecast the next halving with some interval of certainty?

# Data 

```{r}
btc <- read.csv("btc_rewards_history.csv")
```

Looking at all blocks from Block 1 (Genesis, Jan 9th 2009) to Block 826,908 (The last block 
on Jan 22 2024 23:58:49 UTC), including the block reward and the transaction fees received by the miner of the block.


```{r}
reactable(
  btc[1:10, c("BLOCK_TIMESTAMP", "BLOCK_NUMBER", "BLOCK_REWARD", "FEES")]
)
```

We can identify halvings as 50% cuts to the block reward relative to the previous block reward (lagged block reward), which we know occur every 
210,000 blocks. Note, blocks 501,726 and 526,591 were aberrant. Technically miners 
can forget to claim their reward (extremely rare) or not claim the full amount (a few satoshis leftover is not unusual).

```{r}

btc <- btc %>% mutate(
  halving = ifelse(BLOCK_REWARD < 0.6*LAGGED_BLOCK_REWARD, TRUE, FALSE)
)

# Go up to the full day of Jan 22nd. This report is being written as of Jan 23.
btc <- btc[1:826908, ]

# go ahead and label genesis as a halving for simplicity 
btc[1, "halving"] <- TRUE

# two aberrant blocks where the miner incorrectly claimed the amount they could have
btc[btc$BLOCK_NUMBER %in% c('501726', '526591'), "halving"] <- FALSE

reactable(
  btc[btc$halving, c("BLOCK_TIMESTAMP", "BLOCK_NUMBER", "BLOCK_REWARD")]
)

```

# Methods 

Knowing the block number of the next halving (840,000), there are 13,092 
blocks left between Jan 23rd, 2024 12:00AM UTC and the halving.

Aggregating Bitcoin block production to the daily level we can calculate the 
number of blocks per day, the transaction fees earned in that day, and whether the 
day is a halving.

```{r}
btc_daily <- btc %>% mutate(day_ = as.Date(btc$BLOCK_TIMESTAMP)) %>% 
  group_by(day_) %>% summarise(
    n_blocks = n(), 
    total_fees = sum(FEES),
    halving_day = (mean(halving) > 0)
  ) %>% 
  mutate(
     cumulative_blocks = cumsum(n_blocks)
  )

# When the cumulative blocks cycles past a mod 2016, difficulty was adjusted
btc_daily$difficulty_adjusted <- c(FALSE, diff(btc_daily$cumulative_blocks %% 2016) < 0)

reactable(
  tail(btc_daily)
)
```

## Are Blocks Autoregressive?

Every 2,016 blocks (~2 weeks) the difficulty adjustment kicks in. If the previous 
2,016 blocks were produced faster than expected, difficulty goes up to decrease production. If the blocks were produced slower than expected, difficulty falls to increase production.

```{r}

btc_daily_365 <- tail(btc_daily, 365)

p <- plot_ly() %>% 
    add_trace(data = btc_daily_365, 
              x = ~day_, y = ~n_blocks, type = 'scatter', mode = 'markers+lines')

for(i in which(btc_daily_365$difficulty_adjusted)){
  p <- p %>% add_lines(x = btc_daily_365$day_[i], y = c(100, 200), 
              line = list(color = 'lightgray', width = 1, type = 'dash'))
}  
    
p %>% 
  layout(
    title = list(text = "Daily # Blocks over Last 365 Days\n Difficulty Adjustments noted in gray",
                 y = 0.95),
    xaxis = list(title = "Date", showgrid = FALSE),
    yaxis = list(title = "# Blocks Created", showgrid = FALSE),
    showlegend = FALSE,
    grid = FALSE
  )

```

## Are more Blocks made when Fees are higher?

It's rational for more miners to participate in block production when 
transaction fees are high, but given both the difficulty adjustment and how dominant 
block rewards are for miner revenue relative to transaction fees, there is essentially 
little to no correlation between the two over the last 365 days.

```{r}
plot_ly() %>% 
    add_trace(data = tail(btc_daily, 365), 
              x = ~total_fees, y = ~n_blocks, type = 'scatter', mode = 'markers') %>% 
  layout(
    title = list(text = "Blocks Produced Relative to Fees over Last 365 Days", y = 0.975),
    xaxis = list(title = "Fees Paid"),
    yaxis = list(title = "# Blocks Created")
  )

```

# Forecasts 


