---
title: "Predicting the BTC halving"
author: "Charliemarketplace"
date: "`r Sys.Date()`"
output:
  html_document:
    css: "styles.css"
    includes:
      in_header: header.html
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
library(reactable)
library(dplyr)
library(plotly)
library(forecast)
```

# Concept 

The Bitcoin halving occurs every 210,000 blocks. A Bitcoin block occurs *roughly* 
every 10 minutes, but can vary based on hashrate and difficulty adjustments. 

Using just on chain data, can we forecast the next halving with some interval of certainty?

# Data 

```{r}
btc <- read.csv("btc_rewards_history.csv")
```

Looking at all blocks from Block 1 (Genesis, Jan 9th 2009) to Block 826,908 (The last block 
on Jan 22 2024 23:58:49 UTC), including the block reward and the transaction fees received by the miner of the block.


```{r}
reactable(
  btc[1:10, c("BLOCK_TIMESTAMP", "BLOCK_NUMBER", "BLOCK_REWARD", "FEES")]
)
```

We can identify halvings as 50% cuts to the block reward relative to the previous block reward (lagged block reward), which we know occur every 
210,000 blocks. Note, blocks 501,726 and 526,591 were aberrant. Technically miners 
can forget to claim their reward (extremely rare) or not claim the full amount (a few satoshis leftover is not unusual).

```{r}

btc <- btc %>% mutate(
  halving = ifelse(BLOCK_REWARD < 0.6*LAGGED_BLOCK_REWARD, TRUE, FALSE)
)

# Go up to the full day of Jan 22nd. This report is being written as of Jan 23.
btc <- btc[1:826908, ]

# go ahead and label genesis as a halving for simplicity 
btc[1, "halving"] <- TRUE

# two aberrant blocks where the miner incorrectly claimed the amount they could have
btc[btc$BLOCK_NUMBER %in% c('501726', '526591'), "halving"] <- FALSE

reactable(
  btc[btc$halving, c("BLOCK_TIMESTAMP", "BLOCK_NUMBER", "BLOCK_REWARD")]
)

```

# Methods 

Knowing the block number of the next halving (840,000), there are 13,092 
blocks left between Jan 23rd, 2024 12:00AM UTC and the halving.

Aggregating Bitcoin block production to the daily level we can calculate the 
number of blocks per day, the transaction fees earned in that day, and whether the 
day is a halving.

```{r}
btc_daily <- btc %>% mutate(day_ = as.Date(btc$BLOCK_TIMESTAMP)) %>% 
  group_by(day_) %>% summarise(
    n_blocks = n(), 
    total_fees = sum(FEES),
    halving_day = (mean(halving) > 0)
  ) %>% 
  mutate(
     cumulative = cumsum(n_blocks)
  )

# When the cumulative blocks cycles past a mod 2016, difficulty was adjusted
btc_daily$diff_adjusted <- c(FALSE, diff(btc_daily$cumulative %% 2016) < 0)

reactable(
  tail(btc_daily)
)
```

## Are Blocks Autoregressive?

Every 2,016 blocks (~2 weeks) the difficulty adjustment kicks in. If the previous 
2,016 blocks were produced faster than expected, difficulty goes up to decrease production. If the blocks were produced slower than expected, difficulty falls to increase production.

```{r}

btc_daily_365 <- tail(btc_daily, 365)

p <- plot_ly() %>% 
    add_trace(data = btc_daily_365, 
              x = ~day_, y = ~n_blocks, type = 'scatter', mode = 'markers+lines')

for(i in which(btc_daily_365$diff_adjusted)){
  p <- p %>% add_lines(x = btc_daily_365$day_[i], y = c(100, 200), 
              line = list(color = 'lightgray', width = 1, type = 'dash'))
}  
    
p %>% 
  layout(
    title = list(text = "Daily # Blocks over Last 365 Days\n Difficulty Adjustments noted in gray",
                 y = 0.95),
    xaxis = list(title = "Date", showgrid = FALSE),
    yaxis = list(title = "# Blocks Created", showgrid = FALSE),
    showlegend = FALSE,
    grid = FALSE
  )

```

Over the last 365 we see how in between difficulty adjustment periods, more blocks on one 
day tend to correlate with more blocks the next day. This makes sense, as more miners 
can squeeze blocks faster than the expected 2 weeks. 

```{r}

lm_ar1 <- lm(btc_daily_365$n_blocks[2:365] ~ btc_daily_365$n_blocks[1:364])

plot_ly() %>% add_trace(
  data = btc_daily_365, x = ~n_blocks[1:364], y = ~n_blocks[2:365], 
  type = 'scatter', mode = 'markers',
  name = 'N vs N+1'
) %>% 
  add_trace(
    x = ~n_blocks[1:364], 
    y = ~fitted(lm_ar1),
    type = 'scatter',
    mode = 'lines',
    line = list(color = 'pink', dash = 'dash'),
    name = 'LM Fitted'
  ) %>% 
  layout(
    title = list(text = "More blocks predict even more blocks the next day",
                 y = 0.95),
    xaxis = list(title = "# of Blocks on Day N", showgrid = FALSE),
    yaxis = list(title = "# of Blocks on Day N+1", showgrid = FALSE),
    showlegend = FALSE,
    grid = FALSE
  )

```

## Are more Blocks made when Fees are higher?

It's rational for more miners to participate in block production when 
transaction fees are high, but given both the difficulty adjustment and how dominant 
block rewards are for miner revenue relative to transaction fees, there is essentially 
little to no correlation between the two over the last 365 days.

```{r}
plot_ly() %>% 
    add_trace(data = tail(btc_daily, 365), 
              x = ~total_fees, y = ~n_blocks, type = 'scatter', mode = 'markers') %>% 
  layout(
    title = list(text = "Blocks Produced Relative to Fees over Last 365 Days", y = 0.975),
    xaxis = list(title = "Fees Paid"),
    yaxis = list(title = "# Blocks Created")
  )

```

# Forecasts 

The difficulty adjustment works, even knowing the more blocks in one day predict 
more blocks the next day, the adjustment slashes (or skyrockets) production to keep a solid average of 
~150 blocks per day. (147 blocks per day over last 365 days).

A partial autocorrelation of the daily number of blocks over the last 365 days
shows both relationships. The 1 day lag correlation is +0.15 and 15 day lag correction -0.10.

```{r}
pacf(btc_daily_365$n_blocks)
```

## Basic Estimate 

With an average of 147 blocks per day over the last 365 days, and 13,092 blocks between 
Jan 23, 2024 12AM and the Halving, the estimated date could be as simple as: 

`13,092 / 147 = 89 days`

Which would estimate April 21st 2024.

```{r}
as.Date('2024-01-23') + 89
```

## ARIMA simulation

Alternatively, generating an ARIMA model using the known partial autocorrelations, 
we can simulate the next 100 days, 500 times and get the most common date ranges.

```{r}

ar_15 <- arima(btc_daily_365$n_blocks, order = c(15,0,0))

simulations <- data.frame()
p <- plot_ly(type = 'scatter', mode = 'lines')

for(i in 1:500){
run = rep(i, 100)
dates = seq.Date(from = as.Date('2024-01-23'), to = as.Date('2024-05-01'), by = 1)
results = simulate(ar_15, nsim = 100, seed = i)
temp_ = data.frame( run = run, dates = dates, pred = round(as.numeric(results)))
temp_$cumulative = cumsum(temp_$pred)
temp_$halving = 0
# just choosing 200 as a flag in case I want to use as the y value in a chart later
temp_$halving[which(temp_$cumulative >= 13092)[1]] <- 200

simulations = rbind.data.frame(simulations, temp_)

p <- p %>% add_trace(data = temp_, x = ~dates, y = ~pred, type = 'scatter', 
                     mode = 'lines',
                     name = paste0("Run #: ", i))
}

p %>% layout(
  showlegend = FALSE,
  xaxis = list(title = 'Forecast Date'),
  yaxis = list(title = '# Blocks Created in Day'), 
  title = list(text = '500 Simulations of Bitcoin Block Production', y = 0.975)
)
  
```

Doing these 500 simulations, trained on the last 365 days of Bitcoin block data, we can 
identify the most common Halving date as April 20th, 2024. Indicating the basic forecast 
previously made may have not accounted for difficulty adjustment the way this model does. 

```{r}
estimated_dates <- simulations %>% group_by(dates) %>% summarise(
  n = sum(halving == 200)
) %>% filter(n != 0)

plot_ly(estimated_dates, x = ~dates, y = ~n, type = 'bar') %>% layout(
  title = list(text = "Halving Day Frequency across 500 simulations", y = 0.975),
  xaxis = list(title = "Estimated Halving Date"),
  yaxis = list(title = "Simulation Frequency"))

```